Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

### 2024-06-24 optim.tex
edit section 4 and 5

This study explores the potential for a Jellyfish Search Optimiser (JSO)
\citep{chou_novel_2021} -\citep{chou_recent_2022} to be integrated with
projection pursuit guided tour. JSO, inspired by the search behaviour of
jellyfish in the ocean, is a swarm-based metaheuristic designed to solve
global optimisation problems. Compared to traditional methods, JSO
demonstrated stronger search ability and faster convergence, and
requires fewer tuning parameters. These practical advantages make JSO as
a promising candidate for enhancing PP optimisation. As such, the study
aims to explore JSO's potential in enhancing projection pursuit and
examines the characteristics and advantages of such implementation.

@@

\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, breakable, bottomrule=.15mm, toprule=.15mm, opacityback=0, titlerule=0mm, coltitle=black, bottomtitle=1mm, toptitle=1mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, colframe=quarto-callout-note-color-frame, rightrule=.15mm, left=2mm, arc=.35mm, colback=white, opacitybacktitle=0.6, colbacktitle=quarto-callout-note-color!10!white]

@@

The data structure proposed in \citet{RJ-2021-105} is implemented for JS
optimisers and additional information such as index function used, data
dimension, and jellyfish hyper-parameters can be included as additional
columns. An example data collected from finding the sine-wave structure
in 6D data (\texttt{d\ =\ 6}) using a spline index
function(\texttt{splines2d}) is printed below:

@@

The jellyfish hyper-parameters used are 50 jellyfishes
(\texttt{n\_jellies}) and a maximum of 50 tries (\texttt{max\_tries}).
The full data contains 50 simulations, recorded by the variable
\texttt{sim}, with different seeds (\texttt{seed}). Recorded in columns
\texttt{basis}, \texttt{index\_val}, \texttt{tries}, \texttt{loop} and
\texttt{time}, are the sampled projection basis, calculated index value,
iteration ID, jellyfish ID, and time taken to find the optimum,
respectively. The basis column records every visited basis before
comparing with the current (last) basis. If a visited basis has a
smaller index value than the current (last) one, JSO will keep the
current basis for the next iteration in the search but still records the
visited inferior basis. This data can be used to derive numerical
summaries and plots:

\begin{itemize}
\item
  \textbf{Success rate}: The success rate is defined as the proportion
  of simulations that achieve a final index value within 0.05 of the
  best index value found among all 50 simulations. In the dataset above,
  the optimal index value is 0.999. Out of the 50 simulations, 44
  achieved an index value within {[}0.949, 0.999{]}, resulting in a
  success rate of 0.88. Figure~\ref{fig-projection} shows the best
  projection found in each simulation, sorted by the index value. Most
  of the projections show a clear sine-wave structure, while the last
  six show only a straight line with clusters at the end. When local
  optima exist in the projection pursuit problem, plotting the
  projections found in each simulation or by each jellyfish can help
  identify these local optima.
\item
  \textbf{Projection distance}: The projection distance between
  projection bases and the theoretical best basis (or the empirical best
  basis if the theoretical best is unknown) is calculated as the
  Frobenius norm, which measures how close a visited basis is to the
  best basis. Plotting this projection distance against the index value
  for each basis visited reveals how the index value changes as the
  basis approaches the optimal basis, helping to identify the
  squintability of the index function.
  Figure~\ref{fig-index-value-proj-dist} shows the index value against
  the projection distance for the given example and its binned version.
  As the projection distance decreases, the index value initially
  increases with an increasing slope and then levels off. The index
  value shows greater variation at medium projection distances compared
  to small and large distances.
\end{itemize}

@@

\caption{\label{fig-projection}Best projections found by the JS
opitmiser in each simulation, sorted by the index value. The projection
pursuit problem is to detect the 6D sine-wave structure using the spline
index. Most simulations have detected a clear sine-wave pattern, while
the last six show a straight line with clustering at the ends.. In this
simulation, the success rate is 0.88 (44/50).}

@@

The JSO provides a new swarm-based search strategy to optimise
projection pursuit problems. Two examples are provided in this section
to demonstrate its performance. The first example compares its
performance with the search-better optimiser and explores how it behaves
with different combinations of hyper-parameters. The second example
studies the effect of optimisation properties defined in
Section~\ref{sec-theory}, along with jellyfish hyper-parameters, on the
outcome of the optimisation.

@@

Fifty simulations are conducted with both JSO and the search-better , in
four dimensions (\(d = 6, 8, 10, 12\)). The JSO uses 100 jellyfishes and
100 maximum number of tries, while the search-better optimiser allows a
maximum of 1000 samples at each iteration before the algorithm
terminates. Figure~\ref{fig-proj} presents the final projections found
by the two optimisers, broken down by 10th quantile, faceted by the data
dimension. In the 6-dimensional data scenario, JSO consistently
identifies a clear pipe shape. The search-better optimiser also finds
the pipe shape but with a wide rim, suggesting a further polish search
may be required. With increasing dimensions, JSO may not always identify
the pipe shape due to random sampling, but it still presents a pipe
shape in over 50\% of cases. When compared to the search-better
optimiser, JSO reaches higher index values and clearer pipe shapes
across all quantiles.

In the second experiment, fifty simulations are conducted at each
hyper-parameter combination to analyse its effects on the success rate
of JSO . The hyper-parameters tested include: 1) 20, 50, and 100
jellyfishes, and 2) 50 and 100 maximum number of tries. In each
simulation, the success rate is calculated as defined in
Section~\ref{sec-vis}. Figure~\ref{fig-proportion} presents the
proportion of success for each hyper-parameter combinations. As the
number of jellyfishes and maximum tries increase, the success rate also
increases. For simpler problems (6 dimensions), small parameter values
(20 jellyfishes and a maximum number of tries of 50) can already result
in high success rate, while for higher-dimensional problems (8, 10, and
12 dimensions), a combination of 100 jellyfishes and 100 maximum number
of tries is necessary for at least half of the jellyfishes to find the
optimal projection.

@@

The optimisation properties, including smoothness, squintability, and
speed, offer numerical metrics to characterise the complexity of a
projection pursuit optimisation problem. This example investigates how
these metrics, along with the jellyfish search hyper-parameters (the
number of jellyfishes and the maximum number of tries), affect the
success of JSO. Simulations are conducted to obtain the success rate
across various projection pursuit problems, characterised by
shape-to-find, data dimension, and index function used, as well as
different hyper-parameter combinations. Smoothness and squintability are
calculated for each projection pursuit problem as outlined in
Section~\ref{sec-theory}. A generalised linear model is used to
construct the relationship between the success rate, jellyfish
hyper-parameters and optimisation properties.

In addition to the pipe-finding problem, a new shape, sine wave, is
investigated in 6D and 8D spaces with six indices considered:
\texttt{dcor2d\_2}, \texttt{loess2d}, \texttt{MIC}, \texttt{TIC},
\texttt{spline}, and \texttt{stringy}. Combining with two jellyfish
hyper-parameters, a total of 52 cases is produced, comprising of 24
pipe-finding cases and 28 sine-wave finding cases. For each case, JSO is
run fifty times and the success rate is calculated as in
Section~\ref{sec-app-1}.

Smoothness and squintability are computed for each case, following the
procedures outlined in Section~\ref{sec-smoothness} and
Section~\ref{sec-squintability}. To calculate smoothness, three hundred
random bases are simulated. Index values and projection distance (to the
optimal basis) are calculated for each random basis before fitting them
into a Gaussian process model. For squintability, fifty random bases are
sampled and interpolated to the optimal basis with a step size of 0.005.
For these interpolated bases, index values and projection distances to
the optimal basis are calculated and averaged with a bin width of 0.005.
The squintability measure is then calculated from fitting a 4-parameter
scaled logistic function of index value against projection distance,
estimated by non-linear least squares.

Table~\ref{tbl-smoothness-squintability} presents the parameters
estimated from the Gaussian process (variance, range, smooth, and
nugget) and the scaled logistic function (\(\theta_1\) to \(\theta_4\))
for each case. The column ``smooth'' is used as the smoothness measure
and the column ``squint'' is calculated as equation
(\ref{eq-squintability-parametric}) as squintability. The low squint
value of MIC and TIC index is due to its convex shape of the index value
against the projection distance, as shown in
Figure~\ref{fig-idx-proj-dist}. Comparing to other concave shapes, the
maximum first derivative happens at a smaller projection distance
(\(\theta_2\)), requiring the optimiser to get closer to the optimal to
see a significant change in the index value, hence a small squintability
measure.

@@

Table~\ref{tbl-mod-data} combines all the variables calculated above
(the JSO success rate, JSO hyper-parameters, and optimisation
properties) into one table. Three preprocessing steps are applied to the
data: 1) scaling the JSO hyper-parameters by a factor of 10 for
interpretation, 2) creating a new binary variable \texttt{long\_time} to
indicate cases with an average run time over 30 seconds, and 3) encoding
the success rate for the stringy case as 0, because upon visual
inspection of the final projection, none of the 50 simulations
identified the sine-wave shape. A generalised linear model with a
binomial family and a logit link function is used to fit the data and
Table~\ref{tbl-mod-output} presents the model outputs. The model
suggests that the JSO success rate is positively associated with the two
JSO hyper-parameters, as well as with the optimisation properties --
smoothness and squintability. Specifically, using 10 more jellyfish and
10 more tries will increase the odd ratio of success by 24.11\% and
11.93\%, respectively. However, being flagged with long runtime and an
increase of data dimension will reduce the success rate by 41.72\% and
53.36\%, respectively. The variable \texttt{squintability} and
\texttt{dimension} are significant, suggesting their importance relative
to JSO hyper-parameters in the optimisation success. Users can increase
the number of jellyfish and the number of tries to improve the success
rate, until JSO exceeds a 30-second runtime.

>>>>

This study explores the potential for a Jellyfish Search Optimiser (JSO)
to be integrated with projection pursuit guided tour. JSO
\citep{chou_novel_2021} -\citep{rajwar_exhaustive_2023}, inspired by the
search behaviour of jellyfish in the ocean, is a swarm-based
metaheuristic designed to solve global optimisation problems. Compared
to traditional methods, JSO demonstrated stronger search ability and
faster convergence, and requires fewer tuning parameters. These
practical advantages make JSO as a promising candidate for enhancing PP
optimisation. As such, the study aims to explore JSO's potential in
enhancing projection pursuit and examines the characteristics and
advantages of such implementation.

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, opacitybacktitle=0.6, title={Algorithm: Jellyfish Optimizer Pseudo Code}, leftrule=.75mm, coltitle=black, titlerule=0mm, breakable, arc=.35mm, toprule=.15mm, left=2mm, bottomrule=.15mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, colback=white, toptitle=1mm, rightrule=.15mm]

@@

The data structure proposed in \citet{RJ-2021-105} is implemented for
JSO. An example data collected from finding the sine-wave structure in
6D data using a spline index function (\texttt{splines2d}) is printed
below:

@@

The JSO hyper-parameters used are 50 jellyfishes (\texttt{n\_jellies})
and a maximum of 50 tries (\texttt{max\_tries}). The full data contains
50 simulations, indexed by the variable \texttt{sim}, with the
associated \texttt{seed}s. Recorded in columns \texttt{basis},
\texttt{index\_val}, \texttt{tries}, \texttt{loop} and \texttt{time},
are the sampled projection basis, index value, iteration ID, jellyfish
ID, and computational time respectively. The basis column records every
visited basis before comparing with the current (last) basis. If a
visited basis has a smaller index value than the current (last) one, JSO
will keep the current basis for the next iteration in the search but
still records the visited inferior basis. This data can be used to
derive numerical summaries and plots:

\begin{itemize}
\item
  \textbf{Success rate}: The success rate is defined as the proportion
  of simulations that achieves a final index value within 0.05 of the
  best index value found among all 50 simulations. In the dataset above,
  the optimal index value is 0.999. Out of the 50 simulations, 44
  achieved an index value within {[}0.949, 0.999{]}, resulting in a
  success rate of 0.88. Figure~\ref{fig-projection} shows the best
  projection found in each simulation, sorted by the index value. Most
  of the projections show a clear sine-wave structure, while the last
  six show only a straight line with clusters at the end. When local
  optima exist in the projection pursuit problem, plotting the
  projections found in each simulation or by each jellyfish can help
  identify these local optima.
\item
  \textbf{Projection distance}: The projection distance between
  projection bases and the optimal best basis (or the empirical best
  basis if the optimal is unknown) measures how close a visited basis is
  to the best basis and it is calculated as the Frobenius norm in
  Equation

  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \tightlist
  \item
    Plotting projection distances against index values for the bases
    visited reveals how the index value changes as the basis approaches
    the optimal basis, helping to calculate the squintability measure.
    Figure~\ref{fig-index-value-proj-dist} shows the index value against
    the projection distance for the given dataset and its binned
    version. As the projection distance decreases, the index value
    initially increases with an increasing slope and then levels off.
    The index value shows greater variation at medium projection
    distances compared to small and large distances.
  \end{enumerate}
\end{itemize}

@@

\caption{\label{fig-projection}Best projections found by the JSO in each
of the 50 simulations, sorted by index values. The projection pursuit
problem is to detect the 6D sine-wave structure using the spline index.
Most simulations have detected a clear sine-wave pattern, while the last
six show a straight line with clustering at the ends. In this
simulation, the success rate is 0.88 (44/50).}

@@

The JSO provides a new swarm-based search strategy to optimise
projection pursuit problems. Two examples are provided in this section
to demonstrate its performance. The first example compares the
performance of JSO with an exist optimiser, search better, and explores
how JSO behaves with different combinations of hyper-parameters. The
second example studies the effect of optimisation properties defined in
Section~\ref{sec-theory}, along with JSO hyper-parameters, on the
outcome (success rate) of the optimisation.

@@

Fifty simulations are conducted with both JSO and the search better
optimal, in four dimensions (\(d = 6, 8, 10, 12\)). JSO uses 100
jellyfishes and 100 maximum number of tries, while the search-better
optimiser allows a maximum of 1000 samples at each iteration before the
algorithm terminates. Figure~\ref{fig-proj} presents the final
projections found by the two optimisers, broken down by 10th quantile,
faceted by the data dimension. In the 6-dimensional data scenario, JSO
consistently identifies a clear pipe shape. The search better optimiser
also finds the pipe shape but with a wide rim, suggesting a further
polish search may be required. With increasing dimensions, JSO may not
always identify the pipe shape due to random sampling, but it still
presents a pipe shape in over 50\% of cases. When compared to the search
better optimiser, JSO reaches higher index values and clearer pipe
shapes across all quantiles.

In the second experiment, fifty simulations are conducted at each
hyper-parameter combination to analyse its effects on the JSO success
rate. The hyper-parameters tested include: 1) 20, 50, and 100
jellyfishes, and 2) 50 and 100 maximum number of tries. In each
simulation, the success rate is calculated as defined in
Section~\ref{sec-vis}. Figure~\ref{fig-proportion} presents the
proportion of success for each hyper-parameter combinations. As the
number of jellyfishes and maximum tries increase, the success rate also
increases. For simpler problems (6 dimensions), small parameter values
(20 jellyfishes and a maximum number of tries of 50) can already result
in high success rate, while for higher-dimensional problems (8, 10, and
12 dimensions), a combination of 100 jellyfishes and 100 maximum number
of tries is necessary for at least half of the jellyfishes to find the
projection close to the optimal.

@@

The optimisation properties, including smoothness, squintability, and
speed, offer numerical metrics to characterise the complexity of
projection pursuit optimisation problems. This example investigates how
these metrics, along with the hyper-parameters affect the JSO success
rate. Simulations are conducted to obtain the success rate across a
wider array of projection pursuit problems, characterised by
shape-to-find, data dimension, and the index function, as well as
different hyper-parameter combinations. Smoothness and squintability are
calculated for each projection pursuit problem as described in
Section~\ref{sec-theory}. A generalised linear model is used to
construct the relationship between the success rate, JSO
hyper-parameters and optimisation properties.

In addition to the pipe-finding problem, a new shape, sine wave, is
investigated in 6D and 8D spaces with six indexes considered:
\texttt{dcor2d\_2}, \texttt{loess2d}, \texttt{MIC}, \texttt{TIC},
\texttt{spline}, and \texttt{stringy}. Combining with two JSO
hyper-parameters, a total of 52 cases is produced, comprising of 24
pipe-finding cases and 28 sine-wave finding cases. For each case, JSO is
run fifty times and the success rate is calculated as in
Section~\ref{sec-app-1}.

Smoothness and squintability are computed for each case, following the
procedures outlined in Section~\ref{sec-smoothness} and
Section~\ref{sec-squintability}. To calculate smoothness, three hundred
random bases are simulated. Index values and projection distance (to the
optimal basis) are calculated for each random basis before fitting into
a Gaussian process model to obtain the smoothness measure. For
squintability, fifty random bases are sampled and interpolated to the
optimal basis with a step size of 0.005. Index values and projection
distances are calculated for these interpolated bases and the index
values are averaged with a bin width of 0.005. The squintability measure
is then calculated from fitting a 4-parameter scaled logistic function
of index value against projection distance, estimated by non-linear
least squares.

Table~\ref{tbl-smoothness-squintability} presents the parameters
estimated from the Gaussian process (variance, range, smooth, and
nugget) and the scaled logistic function (\(\theta_1\) to \(\theta_4\))
for each case. The column ``smooth'' is used as the smoothness measure
and the column ``squint'' is calculated as equation
\ref{eq-squintability-parametric} as squintability. The low squint value
of MIC and TIC index is due to its convex shape of the index value
against the projection distance, as shown in
Figure~\ref{fig-idx-proj-dist}. Comparing to other concave shapes, the
maximum first derivative happens at a smaller projection distance
(\(\theta_2\)), requiring the optimiser to get closer to the optimal to
see a significant change in the index value, hence a small squintability
measure.

@@

Table~\ref{tbl-mod-data} combines all the variables calculated above
(JSO success rate, hyper-parameters, and optimisation properties) into
one table. Three pre-processing steps are applied to the data: 1)
scaling the JSO hyper-parameters by a factor of 10 for interpretation,
2) creating a new binary variable \texttt{long\_time} to indicate cases
with an average run time over 30 seconds, and 3) encoding the success
rate for the stringy case as 0, because upon visual inspection of the
final projection, none of the 50 simulations identified the sine-wave
shape. A generalised linear model with a binomial family and a logit
link function is used to fit the data and Table~\ref{tbl-mod-output}
presents the model outputs. The model suggests that JSO success rate is
positively associated with the two hyper-parameters, as well as with the
optimisation properties -- smoothness and squintability. Specifically,
using 10 more jellyfishes and 10 more tries will increase the odd ratio
of success by 24.11\% and 11.93\%, respectively. However, being flagged
with long runtime and an increase of data dimension will reduce the
success rate by 41.72\% and 53.36\%, respectively. The variable
\texttt{squintability} and \texttt{dimension} are significant,
suggesting their importance relative to JSO hyper-parameters in the
optimisation success. Users should consider increasing the number of
jellyfish and the number of tries to improve the success rate, until JSO
exceeds a 30-second runtime.
